Categorical Foundations of E-Graphs With Binding

### 2025-07-08 POPL2026/paper/sections/categorical_semantics.tex
comparison update
wip

\section{Categorical semantics of E-Graphs with Bindings}

@@

>>>>

\section{Categorical semantics of E-Graphs with Bindings}
\label{sec:categorical}

@@

\begin{remark}

The above distributivity law closely correspond to one of the congruences maintained by a slotted e-graph, namely

\[
\text{bind}\;s_1\; x \cong \text{bind}\;s_2\;y \;\; \text{if}\; x \cong_{\alpha} y
\]
The congruence intuitively says that if $x$ and $y$ are congruent (up to $\alpha$-equivalence), then the enclosing binders are also congruent.
While this is a \textit{natural} congruence to define when constructing an e-graph with support for binders, in our framework it follows from the free enrichment by a purely categorical argument justifying that it is the \textit{right} congruence to define.
\end{remark}

### 2025-07-08 POPL2026/paper/sections/introduction.tex
wip

Using string diagrams, $\beta$-reduction is simply the re-wiring of the diagram where an application node meets an abstraction, and explicit substitution simplifies into mere composition. 
This is appears to be simpler just by comparing diagrams, but the technical advantages of using string diagrams run deep.
The explicit occurrences of variable names in the older approach raise obvious questions of scope, binding and explicit substitution, whereas the use of terms inside the explicit substitution nodes also raises many technical difficulties which makes proving various mathematical properties essentially intractable.
Additionally, explicit substitutions need to be propagated recursively throughout the graph, which adds another layer of technical and conceptual overhead.
Using de Bruijn indices when performing $\beta$-reduction would hardly solve this problem, as it greatly increases the complexity of representing the reduction as a rewrite rule, and special shifting operators need to be introduced on the right-hand side of a rewrite rule, which is a meta-syntactic operation~\cite{koehler2022sketchguided}.
A recent work of~\cite{slotted-egraphs} on so-called \textit{slotted} e-graphs has addressed the sharing issues when representing terms that are equivalent up to renaming of variables (even free ones) by making variables into first-class citizens of an e-graph and made binding a built-in operation for an e-graph.
An example of a slotted e-graph that represents a term $\lambda x . (a + b) + (x + c)$ can be seen in Figure~\ref{fig:slotted}(left).

@@

A notable feature of a slotted e-graph is that every e-class is parameterised by a set of free variables of the terms represented by this e-class.
The references then additionally convey the information about how variables are instantiated.
For example, an e-class containing $+$ that is parameterised by variable names $s_2$ and $s_3$ gets intantiated once with $a,b$ (via $s_4 \mapsto a, s_2 \mapsto s_4$ and $s_5 \mapsto b, s_3 \mapsto s_5$) and once  with $x,c$.
The fact that the variable $x$ is bound is evident from the corresponding e-class being parameterised by only three variables.
If we were to add another e-node such as $e + d$, there would be a reference added to an already existing e-class such that $s_1 \mapsto e$, $s_2 \mapsto d$.
We can achieve the same degree of sharing (in the sense we have the same number of nodes for $+$) using string diagrams by doing a closure conversion of $+$ (Figure~\ref{fig:slotted}(right)).
However, the addition of these parameters and the corresponding alternations in the algorithmics makes the deeper correspondence between string diagrams and slotted e-graphs obscure.
Despite this, there are applications where the use of string diagrams may be more beneficial than that of (slotted) e-graphs, for example in the case of lambda calculus with explicit substitution that we will describe next.
Also, the categorical model that we built can help to explain the congruences maintained by slotted e-graphs which is further elaborated on in Section~\ref{sec:categorical}.

>>>>

Using string diagrams, $\beta$-reduction is simply the re-wiring of the diagram where an application node meets an abstraction, and explicit substitution simplifies into mere composition. 
This is appears to be simpler just by comparing diagrams, but the technical advantages of using string diagrams run deep.
The explicit occurrences of variable names in the older approach raise obvious questions of scope, binding and explicit substitution, whereas the use of terms inside the explicit substitution nodes also raises many technical difficulties which makes proving various mathematical properties essentially intractable.
Additionally, explicit substitutions need to be propagated recursively throughout the graph, which adds another layer of technical and conceptual overhead.
Using de Bruijn indices when performing $\beta$-reduction would hardly solve this problem, as it greatly increases the complexity of representing the reduction as a rewrite rule, and special shifting operators need to be introduced on the right-hand side of a rewrite rule, which is a meta-syntactic operation~\cite{koehler2022sketchguided}.
A recent work of~\cite{slotted-egraphs} on so-called \textit{slotted} e-graphs has addressed the sharing issues when representing terms that are equivalent up to renaming of variables (even free ones) by making variables into first-class citizens of an e-graph and made binding a built-in operation for an e-graph.
They also solved the problem of explicit variable names guaranteeing freshness inside binders.
An example of a slotted e-graph that represents a term $\lambda x . (a + b) + (x + c)$ can be seen in Figure~\ref{fig:slotted}(left).

@@

A notable feature of a slotted e-graph is that every e-class is parameterised by a set of free variables of the terms represented by this e-class.
The references then additionally convey the information about how variables are instantiated.
For example, an e-class containing $+$ that is parameterised by variable names $s_2$ and $s_3$ gets intantiated once with $a,b$ (via $s_4 \mapsto a, s_2 \mapsto s_4$ and $s_5 \mapsto b, s_3 \mapsto s_5$) and once  with $x,c$.
The fact that the variable $x$ is bound is evident from the corresponding e-class being parameterised by only three variables.
If we were to add another e-node such as $e + d$, there would be a reference added to an already existing e-class such that $s_1 \mapsto e$, $s_2 \mapsto d$.
We can achieve the same degree of sharing (in the sense we have the same number of nodes for $+$) using string diagrams by doing a closure conversion of $+$ (Figure~\ref{fig:slotted}(right)).
However, the addition of these parameters and the corresponding alternations in the algorithmics makes the deeper correspondence between string diagrams and slotted e-graphs obscure.
Despite this, there are applications where the use of string diagrams may be more beneficial than that of (slotted) e-graphs, for example in the case of lambda calculus with explicit substitution that we will describe next \textcolor{red}{Or in Applications section?}.
Also, the categorical model that we built can help to explain the congruences maintained by slotted e-graphs which is further elaborated on in Section~\ref{sec:categorical}.
